A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-05-11 podc_paper/algorithm-GC.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

The space-bounded implementation uses two additional shared arrays:
the \var{leaf} array allows processes to access one another's leaves to perform helping, and
the \var{last} array is used to determine which blocks are safe to discard.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the root  that process $k$}\\
\mbox{ }\Comment{has observed to contain either an enqueue whose value has}\\
\mbox{ }\Comment{been dequeued or a null dequeue }
\end{itemize}

@@

Blocks no longer require the \fld{super} field.  It was used to quickly find a block's 
superblock in the parent node's \fld{blocks} array, but this can now be done efficiently 
by searching the parent's \fld{blocks} RBT instead.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}

>>>>

The space-bounded implementation uses two additional shared arrays:
the \var{leaf} array allows processes to access one another's leaves to perform helping, and
the \var{last} array is used to determine which blocks are safe to discard.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the}\\
\mbox{ }\Comment{root that process $k$ has observed to contain either an en-}\\
\mbox{ }\Comment{queue whose value has been dequeued or a null dequeue }
\end{itemize}

@@

Blocks no longer require the \fld{super} field.  It was used to quickly find a block's 
superblock in the parent node's \fld{blocks} array, but this can now be done efficiently 
by searching the parent's \fld{blocks} RBT instead.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field that is used only for blocks that store a dequeue operation.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the dequeue in the block}
\end{itemize}

### 2023-05-11 podc_paper/boundedspace.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree operations.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a block, just as the \op{CAS} on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful \op{CAS}.
Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other block in the RBT.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
RBT are performed by making a new copy of each RBT node $x$ that must be modified, as well
as each RBT node along the path from the RBT's root to~$x$.
The RBT reachable from the new copy of the root is the result of applying the RBT operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$\var{v}$ in the \ordering\ tree, 
it uses a CAS to swing $\var{v}$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent RBTs in a similar way for a snapshot data structure.

>>>>

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree operations.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a block, just as the \op{CAS} on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful \op{CAS}.
Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other block in the RBT.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
RBT make a new copy of each RBT node $x$ that must be modified, as well
as each RBT node along the path from the RBT's root to~$x$.
The RBT reachable from the new copy of the root is the result of applying the RBT operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$\var{v}$ in the \ordering\ tree, 
it uses a CAS to swing $\var{v}$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent RBTs in a similar way for a snapshot data structure.

### 2023-05-11 podc_paper/correctness-details.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these enqueues is $\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

\superRelationRes*
\begin{proof}
We first show $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs the superblock of $B$ 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
By \Cref{lem::headPosition}, $B.\fld{super} \neq \nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
By \Cref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By \Cref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

@@

\sizeCorrectRes*
\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivial, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By \Cref{lem::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
are the number of enqueues and dequeues contained in $B$.
Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by $L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
\end{proof}

>>>>

Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in\linebreak
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and those in \linebreak
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these enqueues is \linebreak
$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, \linebreak
which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

\superRelationRes*
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
By \Cref{lem::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least 1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
By \Cref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By \Cref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

@@

\sizeCorrectRes*
\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By \Cref{lem::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
are the number of enqueues and dequeues contained in $B$.
Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by $L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
\end{proof}

### 2023-05-11 podc_paper/introduction.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
packet processing \cite{DPDK}, synchronization \cite{KAE23},\here{is this a good citation for this?}
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations  in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show  it is, in fact, possible to design a linearizable, wait-free queue
with step complexity sublinear in $p$.
Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
dequeues take $O(\log^2 p + \log q)$ steps,
and then modify it to bound the space
while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
Both versions use single-word CAS on %reasonably-sized 
small words.
Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
Each process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
Explicitly storing all operations in the tree nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
A preliminary version of this work appeared in~\cite{Nad22}.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

There has been a great deal of research in the past several decades on the design of shared queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management \cite{BBFRSW21}\here{try to find an older, more canonical reference},
packet processing \cite{DPDK}, synchronization \cite{KAE23},\here{is this a good citation for this?}
and sharing resources or tasks.
We focus on shared queues that are \emph{linearizable} \cite{HW90}, meaning that operations
appear to take place atomically, and \emph{lock-free}, meaning that some operation on the queue
is guaranteed to complete regardless of how asynchronous processes are scheduled to take~steps.

The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations  in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show  it is, in fact, possible for a linearizable queue
to have step complexity sublinear in $p$.
Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
It is \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
dequeues take $O(\log^2 p + \log q)$ steps,
and then modify it to bound the space
while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
Both versions of our queue use single-word CAS on reasonably-sized 
words.
We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored in $O(1)$ words.
This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
A process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
Explicitly storing  operations in the tree nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
A preliminary version of this work appeared in~\cite{Nad22}.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-05-11 podc_paper/related.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

\paragraph{List-based Queues.}
The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
See the paper that introduced it for a survey of the early history of concurrent queues.
As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
Thus, it is lock-free but not wait-free and has an amortized step complexity of $\Theta(p)$ per operation.

@@

Many papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty.
% or when concurrent operations can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduce contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
Enqueues within a basket determine their order among themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} described how to use futures to improve
the performance of the MS-queue.
Operations return future
objects instead of responses. Later, when an operation's response is needed, it
can be evaluated using the corresponding future object.
This allows batches of enqueues or dequeues to be done at once on an MS-queue.
However, the implementation satisfies a weaker correctness condition than linearizability.
Milman-Sela et al.~\cite{MKLLP22} extended this approach to allow batches
to mix enqueues and dequeues.
\here{Is this worth saying:
They use some properties of the queue size before and after a batch, similar to a part of our work.}
In the worst case, where operations require their response right away,
batches have size 1, and both of these implementations behave like a standard MS-queue.

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
presented an optimistic  queue implementation. 
%MS-queue uses
%two \nf{CAS}es to do an enqueue: one to change the tail to the new
%node and another one to change the next pointer of the previous node
%to the new node. 
In the MS-queue, an enqueue requires two CAS steps.
The optimistic queue uses a doubly-linked list to reduce the number of
\texttt{CAS} instructions to one in the best case. 
Pointers in the doubly-linked list can be inconsistent, but are fixed when necessary by traversing the list.
Although this fixing is rare in practice, it yields an amortized complexity of $\Omega(qp)$ 
steps per operation for worst-case executions.

@@

\paragraph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free polylogarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue implementation 
that achieves $O(\log p)$ steps for enqueue operations but fails to achieve polylogarithmic step complexity for dequeues: their dequeue operations take $O(k \log p)$ steps if there are $k$ dequeuers.

@@

\paragraph{Other Primitives.}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.

>>>>

\paragraph{List-based Queues.}
The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
The standard Java Concurrency Package includes a version of it.  %java.util.concurrent.ConcurrentLinkedQueue
See \cite{MS98} for a survey of the early history of concurrent queues.
As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
Thus, it is lock-free but not wait-free and has an amortized step complexity of $\Theta(p)$ per operation.

@@

Many papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty.
% or when concurrent operations can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduces contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
Enqueues within a basket order  themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} improved the MS-queue's performance using \emph{futures}.
Operations return future
objects instead of responses. Later, when an operation's response is needed, it
is evaluated using the future object.
This allows batches of enqueues or dequeues to be done at once on an MS-queue.
However, the implementation satisfies a weaker correctness condition than linearizability.
Milman-Sela et al.~\cite{MKLLP22} extended this approach to allow batches
to mix enqueues and dequeues.
\here{Is this worth saying:
They use some properties of the queue size before and after a batch, similar to a part of our work.}
In the worst case, where operations require their response right away,
batches have size 1, and both of these implementations behave like a standard MS-queue.

In the MS-queue, an enqueue requires two CAS steps.
Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
presented an optimistic  queue implementation
%MS-queue uses
%two \nf{CAS}es to do an enqueue: one to change the tail to the new
%node and another one to change the next pointer of the previous node
%to the new node. 
that uses a doubly-linked list to reduce the number of
\texttt{CAS} instructions to one in the best case. 
Pointers in the doubly-linked list can be inconsistent, but are fixed when necessary by traversing the list.
This fixing is rare in practice, but it yields an amortized complexity of $\Omega(qp)$ 
steps per operation in the worst case.

@@

\paragraph{Restricted Queues.}
David gave the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free polylogarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but only for a single dequeuer. 
Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue  
that achieves $O(\log p)$ steps for enqueue operations but fails to achieve polylogarithmic step complexity for dequeues: their dequeue operations take $O(k \log p)$ steps if there are $k$ dequeuers.

@@

\paragraph{Other Primitives.}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue 
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as %particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.
